\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-01 paper/ms.tex

single \celeriteterm\ term with $N=3$ and $c=d=1$:

@@

K &=& \\
&&\begin{pmatrix}

@@

    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-\tau_{2,1}}\left(a\,\cos{\tau_{2,1}}+b\,\sin{\tau_{2,1}}\right)} & {\scriptscriptstyle e^{-\tau_{3,1}}\left(a\,\cos{\tau_{3,1}}+b\,\sin{\tau_{3,1}}\right)}\\ 
    {\scriptscriptstyle e^{-\tau_{2,1}}\left(a\,\cos{\tau_{2,1}}+b\,\sin{\tau_{2,1}}\right)}&{\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)}\\
    {\scriptscriptstyle e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)}&{\scriptscriptstyle  e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)} & \scriptscriptstyle{a + {\sigma_3}^2}

@@

    {\scriptscriptstyle a+{\sigma_1}^2} & {\scriptscriptstyle 0} & {\scriptscriptstyle 0} \\
    {\scriptscriptstyle 0} & {\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle 0} \\
    {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & {\scriptscriptstyle a + {\sigma_3}^2}

@@

    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}

@@

 {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}

@@

Na\"ive implementation of the algorithm in \eq{algorithm} will have numerical
overflow and underflow issues in many realistic cases because both $U$ and $V$
have factors of the form $e^{\pm t}$, where $t$ can be any arbitrarily large
real number.
This can be avoided with the appropriate reparameterization.
We define the following pre-conditioned variables
\begin{eqnarray}
\tilde{U}_{n,2\,j-1} &=& a_j\,\cos(d_j\,t_n) + b_j\,\sin(d_j\,t_n) \\
\tilde{U}_{n,2\,j} &=& a_j\,\sin(d_j\,t_n) - b_j\,\cos(d_j\,t_n) \\
\tilde{V}_{m,2\,j-1} &=& \cos(d_j\,t_m) \\
\tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \\
\tilde{X}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,X_{n,2\,j-1} \\
\tilde{X}_{n,2\,j} &=& e^{-c_j\,t_n}\,X_{n,2\,j}
\end{eqnarray}
and the set of variables
\begin{eqnarray}
\phi_{n,2\,j-1} = \phi_{n,2\,j} = e^{-c_j\,(t_n - t_{n-1})}
\end{eqnarray}
with the constraint
\begin{eqnarray}
\phi_{1,2\,j-1} = \phi_{1,2\,j} = 0 \quad.
\end{eqnarray}
Using this parameterization, we find the following, numerically stable,
algorithm
\begin{eqnarray}
    S_{n,j,k} &=& \phi_{n,j}\,\phi_{n,k}\,\left[S_{n-1,j,k} +
    D_{n-1,n-1}\,\tilde{X}_{n-1,j}\,\tilde{X}_{n-1,k}\right] \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} \tilde{U}_{n,j}\,S_{n,j,k}\,\tilde{U}_{n,k}
    \nonumber\\
\tilde{X}_{n,j} &=& \frac{1}{D_{n,n}}\left[ \tilde{V}_{n,j} -
    \sum_{k=1}^{2\,J}\tilde{U}_{n,k}\,S_{n,j,k} \right] \quad.
\end{eqnarray}

@@

>>>>

single \celeriteterm\ term with $c=d=1$: %and $N=3$ 

@@

K&=&%

@@

%    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-\tau_{2,1}}\left(a\,\cos{\tau_{2,1}}+b\,\sin{\tau_{2,1}}\right)} & {\scriptscriptstyle e^{-\tau_{3,1}}\left(a\,\cos{\tau_{3,1}}+b\,\sin{\tau_{3,1}}\right)}\\ 
%    {\scriptscriptstyle e^{-\tau_{2,1}}\left(a\,\cos{\tau_{2,1}}+b\,\sin{\tau_{2,1}}\right)}&{\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)}\\
%    {\scriptscriptstyle e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)}&{\scriptscriptstyle  e^{-\tau_{3,2}}\left(a\,\cos{\tau_{3,2}}+b\,\sin{\tau_{3,2}}\right)} & \scriptscriptstyle{a + {\sigma_3}^2}
    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-\tau_{21}}\left(a\,\cos{\tau_{21}}+b\,\sin{\tau_{21}}\right)} & {\scriptscriptstyle e^{-\tau_{31}}\left(a\,\cos{\tau_{31}}+b\,\sin{\tau_{31}}\right)} & \ldots\\ 
    {\scriptscriptstyle e^{-\tau_{21}}\left(a\,\cos{\tau_{21}}+b\,\sin{\tau_{21}}\right)}&{\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)} & \ldots\\
    {\scriptscriptstyle e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)}&{\scriptscriptstyle  e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)} & \scriptscriptstyle{a + {\sigma_3}^2} & \ldots \\
    \ldots & \ldots & \ldots & \ldots

@@

%    {\scriptscriptstyle a+{\sigma_1}^2} & {\scriptscriptstyle 0} & {\scriptscriptstyle 0} \\
%    {\scriptscriptstyle 0} & {\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle 0} \\
%    {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & {\scriptscriptstyle a + {\sigma_3}^2}
    {\scriptscriptstyle a+{\sigma_1}^2} & {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & \ldots\\
    {\scriptscriptstyle 0} & {\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle 0} & \ldots\\
    {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & {\scriptscriptstyle a + {\sigma_3}^2} & \ldots\\
%    {\scriptscriptstyle a+{\sigma_1}^2} & {\scriptscriptstyle 0} & \ldots \\
%    {\scriptscriptstyle 0} & {\scriptscriptstyle a+{\sigma_2}^2 }& \ldots \\
    \ldots & \ldots & \ldots & \ldots

@@

    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}\\
   \ldots & \ldots

@@

 {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}\\
 \ldots & \ldots

@@

Na\"ive implementation of the algorithm in \eq{algorithm} will have numerical
overflow and underflow issues in many realistic cases because both $U$ and $V$
have factors of the form $e^{\pm c_j t}$, where $t$ can be any arbitrarily large
real number.
This can be avoided with the appropriate reparameterization.
We define the following pre-conditioned variables
\begin{eqnarray}
\tilde{U}_{n,2\,j-1} &=& a_j\,\cos(d_j\,t_n) + b_j\,\sin(d_j\,t_n) \\
\tilde{U}_{n,2\,j} &=& a_j\,\sin(d_j\,t_n) - b_j\,\cos(d_j\,t_n) \\
\tilde{V}_{m,2\,j-1} &=& \cos(d_j\,t_m) \\
\tilde{V}_{m,2\,j} &=& \sin(d_j\,t_m) \\
\tilde{X}_{n,2\,j-1} &=& e^{-c_j\,t_n}\,X_{n,2\,j-1} \\
\tilde{X}_{n,2\,j} &=& e^{-c_j\,t_n}\,X_{n,2\,j}
\end{eqnarray}
and the set of variables
\begin{eqnarray}
\phi_{n,2\,j-1} = \phi_{n,2\,j} = e^{-c_j\,(t_n - t_{n-1})}
\end{eqnarray}
with the constraint
\begin{eqnarray}
\phi_{1,2\,j-1} = \phi_{1,2\,j} = 0 \quad.
\end{eqnarray}
Using this parameterization, we find the following, numerically stable,
algorithm
\begin{eqnarray}
    S_{n,j,k} &=& \phi_{n,j}\,\phi_{n,k}\,\left[S_{n-1,j,k} +
    D_{n-1,n-1}\,\tilde{X}_{n-1,j}\,\tilde{X}_{n-1,k}\right] \nonumber\\
D_{n,n} &=& A_{n,n} -
    \sum_{j=1}^{2\,J}\sum_{k=1}^{2\,J} \tilde{U}_{n,j}\,S_{n,j,k}\,\tilde{U}_{n,k}
    \nonumber\\
\tilde{X}_{n,j} &=& \frac{1}{D_{n,n}}\left[ \tilde{V}_{n,j} -
    \sum_{k=1}^{2\,J}\tilde{U}_{n,k}\,S_{n,j,k} \right] \quad.
\end{eqnarray}

@@

Simulation of a Gaussian process for large $N$ requires multiplying an
$N$-dimensional vector of Gaussian normal deviates, $\bf{q}$, by the square root
of the covariance matrix:
\begin{equation}
{\bf y} = K_\bvec{\alpha}^{1/2} {\bf q},
\end{equation}
where $q_i \sim N(0,1)$.  Note that $K_\bvec{\alpha} = K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{1/2}\right)^T$, which
comparing with the Cholesky factorization gives $K_\bvec{\alpha}^{1/2} = L D^{1/2}$,
where $(D^{1/2})_{ii} = \sqrt{D_{ii}}$.  Also note that the exponent of
the likelihood with $\mu_\bvec{\theta} =0$ is given by
\begin{eqnarray}
Q &=&-\frac{1}{2}{\bf y}^T K_\bvec{\alpha}^{-1} {\bf y}\\
&=& -\frac{1}{2}{\bf y}^T \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} {\bf y}\\
&=& -\frac{1}{2}{\bf q}^T K_\bvec{\alpha}^{1/2} \left(K_\bvec{\alpha}^{-1/2}\right)^T K_\bvec{\alpha}^{-1/2} K_\bvec{\alpha}^{1/2} {\bf q}\\
&=& -\frac{1}{2}{\bf q}^T {\bf q}.
\end{eqnarray}
The last quantity is simply the dot product of an $N$-dimensional vector of independent standard normal random variables with itself,
which is distributed according to the chi-squared distribution with $N$ degrees of freedom, i.e.\ $Q \sim \chi_N^2$.

To generate a draw from a Gaussian process with a specified \celerite\ model requires carrying out the
low-rank Cholesky decomposition first, generating the vector ${\bf q}$ of random normal deviates, and then multiplying
\begin{equation}
{\bf y} = L D^{1/2} {\bf q}.
\end{equation}
This may be carried out in $O(JN)$ operations with
\begin{eqnarray}
f_{n,j} &=& \phi_{n-1,j} \left(f_{n-1,j} + \tilde{X}_{n-1,j} q_{n-1}\right)\\
y_n &=& D_{n,n}^{1/2} q_n + \sum_{k=1}^{2J} \tilde{U}_{n,k} f_{n,k}
\end{eqnarray}
for $n = 1$ to $N$ and $j=1$ to $2J$, with $f_{0,j} = 0$ for all $j$ and $q_0 = 0$.
\agoltodo{Check that these equations are in fact correct.}
We expect that this simulation algorithm will be useful for test-driving Gaussian
process inference algorithms applied to large datasets; even simulation of
Gaussian processes becomes prohibitive for large $N$ if standard Cholesky decomposition 
and multiplication are used.

Multiplication of the full covariance matrix times a vector can be accomplished
by yet another multiplication by the transpose of the Cholesky decomposition.
\dfmtodo{Should we spell this out as well?}

\subsection{Prediction}

\agoltodo{Write up the prediction algorithm.}
